---
title: "interfacer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{interfacer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(interfacer)
```

# Defining an interface

An interface defines the structure of acceptable dataframes. It is a list
of column names, plus types and some documentation about the column.

```{r}
i_test = iface(
  id = integer ~ "an integer ID",
  test = logical ~ "the test result"
)
```

Printing an interface specification shows the structure that the `iface` defines.

```{r, results='markup'}
cat(print(i_test))
```

An interface specification is enforced as the default value for a function. This
is a dummy default value but when combined with `ivalidate` in the function body
a passed parameter can be validated to ensure the dataframe is of the right
shape. We can use `iparam` in the Roxygen skeleton to document the dataframe
constraints.

```{r}
#' An example function
#'
#' @iparam mydata a dataframe
#' @param another an example   
#' @param ... not used
#'
#' @return the conformant dataframe
#' @export
example_fn = function(
  mydata = i_test,
  another = "value",
  ...
) {
  mydata = ivalidate(mydata, ...)
  return(mydata)
}
```

In this case when we later call `example_fn` the data is checked against the 
requirements by `ivalidate`, and if acceptable passed on to the rest of the function body (in this case the result is returned).

```{r}
example_data = tibble::tibble(
    id=c(1,2,3),
    test = c(TRUE,FALSE,TRUE)
  )

# this returns the qualifying data
example_fn(
  example_data, 
  "value for another"
) %>% dplyr::glimpse()
```

If we pass non-conformant data `ivalidate` throws an informative error about
what is wrond with the data. In this case the `test` column is missing:

```{r}
bad_example_data = tibble::tibble(
    id=c(1,2,3),
    wrong_name = c(TRUE,FALSE,TRUE)
  )

# this causes an error as example_data_2$wrong_test is wrongly named
try(example_fn(
  bad_example_data, 
  "value for another"
))
```

We can recover from this error by renaming the columns before passing
`bad_example_data` to `example_fn()`. Alternatively we can use the dynamic dots
to pass a mapping to `ivalidate` which is applied before the validation. This
parameter is called `.imap`, and its contents is a `imapper()`  that holds an
set of expressions used for a `dplyr::mutate` call).

```{r}
example_fn(
  bad_example_data, 
  "value for another",
  .imap = interfacer::imapper(test = wrong_name)
) %>% dplyr::glimpse()

```


# Extension and composition

To be useful interface specifications can be composed and extended. In this case
a subtype of the `iface` specification can be created:

```{r}
i_test_extn = iface(
  i_test,
  extra = character ~ "a new value",
  .groups = FALSE
)

print(i_test_extn)
```

and then this extended `iface` used to constrain the input of another function
by specifying it as the default value. We can also constrain the output of the
function to be conformant to another specification usng `ireturn`. Examples of
documenting the input parameter and the output parameter are provided here:

```{r}

#' Another example function 
#' 
#' @param mydata `r idocument(example_fn2, mydata)`
#' @param another an example   
#' @param ... not used
#'
#' @return `r i_test`
#' @export
example_fn2 = function(
    mydata = i_test_extn,
    ...
) {
  mydata = ivalidate(mydata, ..., .prune = TRUE)
  mydata = mydata %>% dplyr::select(-extra)
  # check the return value conforms to a new specification
  ireturn(mydata, i_test)
}
```

In this case the `ivalidate` call prunes unneeded data from the dataframe,
removing any extra columns, and also ensures that the input is not grouped in
any way. Grouping

```{r}
grouped_example_data = tibble::tibble(
    id = c(1,2,3),
    test = c(TRUE,FALSE,TRUE),
    extra = c("a","b","c"),
    unneeded = c("x","y","z")
  ) %>% dplyr::group_by(id)
```

This is rejected because the grouping is incorrect. An informative error message
is provided:

```{r}
try(example_fn2(grouped_example_data))
```

Following the instructions in the error message makes this previously failing data validate against `i_test_extn`:

```{r}
grouped_example_data %>% 
  dplyr::ungroup() %>% 
  example_fn2() %>% 
  dplyr::glimpse()
```

# Documentation

The roxygen block of documentation for this second interface is determined by
the `#' @iparam` block and the function `idocument` which takes a function name
and optionally a parameter name and writes an informative block about the
dataframe expected by the function:

```{r}
cat(idocument(example_fn2))
```

It is also possible to use a custom `roxygen2` tag `@iparam <param> <description>`
which will automatically in-line the description of the structure of the dataframe 
into the documentation.

# Type coercion

`interfacer` does not implement a rigid type system, but rather a permissive one. 
If the provided data can be coerced to the specified type without major loss then 
this is automatically done, as long as it can proceed with no warnings. In this 
example `id` (an int) is provided as a `character` and `extra` (a character) is
coerced from the provided numeric.

```{r}
example_fn2(
  tibble::tibble(
    id=c("1","2","3"),
    test = c(TRUE,FALSE,TRUE),
    extra = 1.1
  )) %>% dplyr::glimpse()
```

Completely incorrect data types on the other hand are picked up and rejected. In
this case the data supplied for `id` cannot be cast to integer without loss.
Similar behaviour is seen if logical data is anything other than 0 or 1 for
example.

```{r}
try(example_fn(
  tibble::tibble(
    id= c("1.1","2","3"),
    test = c(TRUE,FALSE,TRUE)
  )))
```

Factors might have allowable levels as well. For this we define them as an
`enum` which accepts a list of values, which then must be matched by the levels
of a provided factor. The order of the levels will be taken from the `iface`
specification and re-levelling of inputs is taken to ensure the factor levels
match the specification. If `.drop = TRUE` is specified then values which don't
match the levels will be cast to `NA` rather than causing failure to allow
conformance to a subset of factor values.

```{r}

if (rlang::is_installed("ggplot2")) {
  
  i_diamonds = iface( 
    color = enum(D,E,F,G,H,I,J,extra) ~ "the colour",
    cut = enum(Ideal, Premium, .drop=TRUE) ~ "the cut",
    price = integer ~ "the price"
  )
  
  # itest(ggplot2::diamonds, i_diamonds)
  iconvert(ggplot2::diamonds, i_diamonds, .prune = TRUE)
   
} %>% dplyr::glimpse()
```


# Grouping

an `iface` specification can permissive or dogmatic about the grouping of the input. If
the .groups option in an `iface` specification is NULL (e.g. `iface(..., .groups=NULL)`)
then any grouping is allowed. If it is FALSE then no grouping is allowed. The
third option is to supply a one sided formula. In this case the variables in the
formula define the grouping that must be present, e.g. `~ grp1 + grp2`, but if
it also includes a `.`, then additional grouping is also permitted (e.g. 
`~ . + grp1 + grp2`) allows a grouping such as `df %>% group_by(anything, grp1, grp2)`.
N.B. ordering of grouping mostly is as expected but is a work in progress, it makes
most sense for mandatory groups to be the right most / most granular groups.

```{r}

i_diamonds = interfacer::iface(
	carat = numeric ~ "the carat column",
	color = enum(`D`,`E`,`F`,`G`,`H`,`I`,`J`, .ordered=TRUE) ~ "the color column",
	x = numeric ~ "the x column",
	y = numeric ~ "the y column",
	z = numeric ~ "the z column",
	.groups = ~ . + carat + cut
)

# permissive grouping with the `~ . + carat + cut` groups rule
ggplot2::diamonds %>% 
  dplyr::group_by(color, carat, cut) %>% 
  iconvert(i_diamonds, .prune = TRUE) %>% 
  dplyr::glimpse()

```

If a group column is specified it must be present, regardless of the rest of the
`iface` specification. So in this example `cut` is required by the `i_diamonds` contract
but it can be any format.

# Dispatch

Dispatching to one of a number of functions based on the nature of a dataframe
input is enabled by `idispatch(...)`. This emulates the behaviour of `S3`
classes but for dataframes, based on their columns and also their grouping. An
example is the following:

```{r}

disp_example = function(x, ...) {
  idispatch(x,
    disp_example.extn = i_test_extn,
    disp_example.no_extn = i_test
  )
}

disp_example.extn = function(x = i_test_extn, ...) {
  # idispatch has already validated x so we don't need to do it again here
  message("extended data")
  return(colnames(x))
}

disp_example.no_extn = function(x = i_test, ...) {
  # in this toy example the function content is very similar.
  message("not extended data")
  return(colnames(x))
}
```

If we call `disp_example()` with data that matches the `i_test_extn` specification
we get one type of behaviour:

```{r}

tmp = tibble::tibble(
    id=c("1","2","3"),
    test = c(TRUE,FALSE,TRUE),
    extra = 1.1
)

tmp %>% disp_example()
```

But if we call `disp_example()` with data that only matches the `i_test` specification
we get different behaviour:

```{r}
# this matches the i_test_extn specifciation:
tmp2 = tibble::tibble(
    id=c("1","2","3"),
    test = c(TRUE,FALSE,TRUE)
)

tmp2 %>% disp_example()
```


## Recursive dispatch

The order of the rules is important with in general the more detailed needing to 
be provided first. We can leverage this to create a recursive functional pattern of 
dataframe processing that allows multiple inputs to converge on a single output, this
also demonstrates the use of `itest()` which simply checks an input conforms to 
an `iface` spec:

```{r}

i_input_1 = iface(
  x = integer ~ "the positives",
  n = default(100) + integer ~ "the total"
)

i_input_2 = iface(
  p = proportion ~ "the proportion",
  n = default(100) + integer ~ "the total"
)

i_interim = iface(
  i_input_1,
  i_input_2
)

i_final = iface(
  i_interim,
  lower = double ~ "wilson lower CI",
  upper = double ~ "wilson lower CI",
  mean = double ~ "wilson mean"
)

i_target = iface(
  i_final,
  label = character ~ "a printable label"
)

process.input_1 = function(x = i_input_1,...) {
  return(x %>% dplyr::mutate(p = x/n))
}

process.input_2 = function(x = i_input_1,...) {
  return(x %>% dplyr::mutate(x = floor(p*n)))
}

process.interim = function(x) {
  return(x %>% dplyr::mutate(binom::binom.wilson(x,n)))
}

process.final = function(x) {
  return(x %>% dplyr::mutate(label = sprintf("%1.1f%% [%1.1f%% - %1.1f%%] (%d/%d)", 
    mean*100, lower*100, upper*100, x, n)))
}

process = function(x,...) {
  # this test must be at the front to prevent infinite recursion
  if (itest(x, i_target)) return(x)
  out = idispatch(x,
    process.final = i_final,
    process.interim = i_interim,
    process.input_2 = i_input_2,
    process.input_1 = i_input_1
  )
  return(process(out))
}

```


Processing an input of `type 1` results in one path through the data pipeline:

```{r}
# tibble::tibble(x=c(10,30), n=c(NA,50)) %>% itest(i_input_1)
process(tibble::tibble(x=c(10,30), n=c(NA,50))) %>% dplyr::glimpse()
```

Processing an input of `type 2`, results in a different path through the data
pipeline, but the same outcome:

```{r}
# tibble::tibble(p=0.15,n=1000) %>% itest(i_input_2)
process(tibble::tibble(p=0.15,n=1000)) %>% dplyr::glimpse()
```
Care must be taken though in this pattern, particularly if you are re-using
column names,as datatype coercion could result in some column types being
switched backwards and forwards, and other infinite loop problems.

## Grouping based dispatch

It is often useful to have a function that can expects a specific grouping but
can handle additional groups. One way of handling these is to use `purrr` and
nest columns extensively. Nesting data in the unexpected groups and repeatedly
applying the function you want. An alternative `dplyr` solution is to use a
`group_modify`. `interfacer` leverages this second option to automatically
determine a grouping necessary for a pipeline function from the stated grouping
requirements and automatically handle them without additional coding in the
package.

For example if we have the following `iface` the input for a function must be 
grouped only by the `color` column:

```{r}
i_diamond_price = interfacer::iface(
  color = enum(`D`,`E`,`F`,`G`,`H`,`I`,`J`, .ordered=TRUE) ~ "the color column",
  price = integer ~ "the price column",
  .groups = ~ color
)
```

A package developer writing a pipeline function may use this fact to handle 
possible additional grouping by using a `igroup_process(df, ...)` 

```{r}
 # exported function in package
 # at param can use `r idocument(ex_mean, df)` for documentation
 ex_mean = function(df = i_diamond_price, extra_param = ".") {

   # dispatch based on groupings:
   igroup_process(df,

     # the real work of this function is provided as an anonymous inner
     # function (but can be any other function e.g. package private function)
     # or a purrr style lambda.

     function(df, extra_param) {
       message(extra_param, appendLF = FALSE)
       return(df %>% dplyr::summarise(mean_price = mean(price)))
     }

   )
 }
```

If we pass this to correctly grouped data conforming to `i_diamond_price` the
inner function is executed once transparently, after the input has been validated:

```{r}
# The correctly grouped dataframe
ggplot2::diamonds %>%
  dplyr::group_by(color) %>%
  ex_mean(extra_param = "without additional groups...") %>%
  dplyr::glimpse()
```

If on the other hand additional groups are present the inner function is 
executed once for each of the additional groups. Data validation happens once
per group, which affects interpretation of uniqueness.

```{r}
# The incorrectly grouped dataframe
ggplot2::diamonds %>%
  dplyr::group_by(cut, color, clarity) %>%
  ex_mean() %>%
  dplyr::glimpse()

```

The output of this is actually grouped by `cut` then `clarity` as
`color` column is consumed by the nested function in `igroup_process`.

# Complex rules & composition

In the section above we used the pseudo-function `default(100) + integer` to 
specify the input column `x`. We also talked about the use of `enum(...)` as a 
rule to match factors. A range of other possible modifiers are possible such as
`group_unique` or `complete` which enforce uniqueness and completeness of factors
or natural numeric sequences.


# Default values

```{r}

i_iris = interfacer::iface(
	Sepal.Length = numeric ~ "the Sepal.Length column",
	Sepal.Width = numeric ~ "the Sepal.Width column",
	Petal.Length = numeric ~ "the Petal.Length column",
	Petal.Width = numeric ~ "the Petal.Width column",
	Species = enum(`setosa`,`versicolor`,`virginica`) ~ "the Species column",
	.groups = NULL,
  .default = TRUE
)

test_fn = function(i = i_iris, ...) {
  i = ivalidate(i,...)
  return(i)
}

# Outputs a zero length data frame
test_fn() %>% dplyr::glimpse()

```


# Nesting & list columns

```{r}

i_diamonds_cat = interfacer::iface(
  cut = enum(`Fair`,`Good`,`Very Good`,`Premium`,`Ideal`, .ordered=TRUE) ~ "the cut column",
  color = enum(`D`,`E`,`F`,`G`,`H`,`I`,`J`, .ordered=TRUE) ~ "the color column",
  clarity = enum(`I1`,`SI2`,`SI1`,`VS2`,`VS1`,`VVS2`,`VVS1`,`IF`, .ordered=TRUE) ~ "the clarity column",
  data = list(i_diamonds_data) ~ "A nested data column must be specified as a list",
  .groups = FALSE
)

i_diamonds_data = interfacer::iface(
  carat = numeric ~ "the carat column",
  depth = numeric ~ "the depth column",
  table = numeric ~ "the table column",
  price = integer ~ "the price column",
  x = numeric ~ "the x column",
  y = numeric ~ "the y column",
  z = numeric ~ "the z column",
  .groups = FALSE
)

nested_diamonds = ggplot2::diamonds %>%
  tidyr::nest(data = c(-cut,-color,-clarity))

nested_diamonds %>% iconvert(i_diamonds_cat) %>% dplyr::glimpse()

```


# Developer tools

TBD: 

* `iclip(diamonds)`
* `use_dataframe(iris)`
* `use_iface(diamonds)`

